Manuscript_Github_Pchabaudi_phaseGene

### 2022-07-28 chapter4_results_chabaudi24h/Thesis_Github_Pchabaudi_phaseGenes.Rmd
Make phaseograms of the most signif GO annotation genes and make graphs nicer

library('readr')
library('dplyr')
library('ggplot2')
library("RColorBrewer")
library('data.table')
library('tibble')
library('stringr')
library('kableExtra')
library('viridis')
library('reshape2')
library('ComplexHeatmap')
library('circlize') # for circular plots and colour
library("factoextra")
library("MatrixGenerics") # for rowMins and rowMax
library("topGO")
library("FoldGO")

@@

```{r times_stages_labels}
#Use the RMT stage differential count in order to name the samples, but keep time to retain order
times_stages_table <- data.frame(time = c('02h', '05h', '08h', '11h', '14h', '17h','20h','23h'),
                                 stage_diff_count = c('Late-Rings',
                                                      'Rings-Troph-conversion',
                                                      'Early-Trophs',
                                                      'Mid-Trophs',
                                                      'Late-Trophs',
                                                      'Early-Rings',
                                                      'Early-Mid-Rings',
                                                      'Late-Mid-Rings'))
# R will convert the dashes to '.' at some point so let's have this notation here also
times_stages_table$r_stage_diff_count <- str_replace_all(times_stages_table$stage_diff_count, 
                                                         pattern = '-', 
                                                         replacement = '\\.')
```

@@

```{r import_z}
chab_z.df <- inner_join(chab_z.list[[1]]$zscores, 
                        chab_z.list[[2]]$zscores, 
                        by = 'Geneid')
chab_z.df <- read.csv(file = 'data/Pchabaudi_z.csv')

chab_z.list <- list('rMT' = list('zscores' = chab_z.df[,grepl(colnames(chab_z.df),
                                                                             pattern = 'Geneid|rMT')]),
                    'SBP' = list('zscores' = chab_z.df[,grepl(colnames(chab_z.df),
                                                                             pattern = 'Geneid|SBP')]))

@@

```{r phase_go_enrich}
#Get the gene list for each group of genes in and out of phase.
gene.list <- chabaudi_gene_info$Geneid
z_max_diff.list <- lapply(seq(0,4), 
                          function(diff_num) {
                            diff.list <- filter(z_max_diff, diff == diff_num) %>% 
                              .$Geneid
                            #Get 1 for on the list and 0 for not.
                            gene_diff.list <- as.integer(gene.list %in% diff.list)
                            names(gene_diff.list) <- gene.list
                            return(gene_diff.list)
                          })
names(z_max_diff.list) <- seq(0,4)

@@

```{r}
#Load for the showSigOfNodes network function.
library('Rgraphviz')

@@

#Get first (most significant) enriched GO id 
goID <- goRes[4, "GO.ID"]
#Show distribution of GO term genes compared to null distribution (all other genes)
print(showGroupDensity(GOdata, goID, ranks = TRUE))

@@

draw_heatmap_zscore(zscore_list = chab_z.list,
                    filter_genes_vector = names(z_max_diff.list$`1`[z_max_diff.list$`2` == 1]),
                    gene_ordering_vector = phaseogram_time_order$rMT$zscore_order$Geneid,
                    threshold_tpm = 1,
                    use_times = TRUE,
                    use_order_vector = FALSE,
                    show_row_names = FALSE,
                    tpm_mat = TRUE)
```

@@

lapply(seq(0,4), 
       function(phase){
         GOdata <- new("topGOdata", 
                       ontology = "BP", 
                       geneSel = function(x){x == 1},
                       allGenes = getElement(z_max_diff.list, paste(phase)),
                       nodeSize = 5,
                       annot = annFUN.gene2GO, 
                       gene2GO = go_annotation)
         resultFis <- runTest(GOdata, 
                              algorithm = "classic", 
                              statistic = "fisher")
         draw_heatmap_zscore(zscore_list = chab_z.list,
                             filter_genes_vector = names(
                               getElement(z_max_diff.list, paste(phase))[
                                 getElement(z_max_diff.list, paste(phase)) == 1
                               ]
                             ),
                             gene_ordering_vector = phaseogram_time_order$rMT$zscore_order$Geneid,
                             threshold_tpm = 1,
                             use_times = TRUE,
                    use_order_vector = FALSE,
                    show_row_names = FALSE,
                    tpm_mat = TRUE)
         GenTable(GOdata, 
                  classic = resultFis, 
                  orderBy = "classic", 
                  ranksOf = "classic",
                  topNodes = 25) %>% 
           filter(classic < 0.05)
       })

@@

>>>>

library('readr')
library('dplyr')
library('ggplot2')
library("RColorBrewer")
library('data.table')
library('tibble')
library('stringr')
library('kableExtra')
library('viridis')
library('reshape2')
library('ComplexHeatmap')
library('circlize') # for circular plots and colour
library("factoextra")
library("MatrixGenerics") # for rowMins and rowMax
library("topGO")
library("FoldGO")
library('Rgraphviz')

@@

```{r times_stages_labels}
#Use the RMT stage differential count in order to name the samples, but keep time to retain order
times_stages_table <- data.frame(time = c('02h', '05h', '08h', '11h', '14h', '17h','20h','23h'),
                                 stage_diff_count = c('Late-Rings',
                                                      'Rings-Troph-conversion',
                                                      'Early-Trophs',
                                                      'Mid-Trophs',
                                                      'Late-Trophs',
                                                      'Early-Rings',
                                                      'Early-Mid-Rings',
                                                      'Late-Mid-Rings'))
# R will convert the dashes to '.' at some point so let's have this notation here also
times_stages_table$r_stage_diff_count <- str_replace_all(times_stages_table$stage_diff_count, 
                                                         pattern = '-', 
                                                         replacement = '\\.')
#Add a number from 1 to 8 for each stage, useful for phase calculations
times_stages_table$num <- 1:8
```

@@

```{r import_z}
chab_z.df <- read.csv(file = 'data/Pchabaudi_z.csv')

#Make list object of the TPM values and the z-scores value by concatenating the 
# lists of the zscore dataframes and the TPM lists from above.
chab_z.list <- list('RMT' = c(chab_tpm_avg.list[['RMT']],
                              list('zscores' = chab_z.df[,grepl(colnames(chab_z.df),
                                                           pattern = 'Geneid|rMT')])),
                    'SBP' = c(chab_tpm_avg.list[['SBP']],
                              list('zscores' = chab_z.df[,grepl(colnames(chab_z.df),
                                                           pattern = 'Geneid|SBP')])))

@@

```{r z_genelist}
#Get the gene list for each group of genes in and out of phase.
gene.list <- chabaudi_gene_info$Geneid
z_max_diff.list <- lapply(seq(0,4), 
                          function(diff_num) {
                            diff.list <- filter(z_max_diff, diff == diff_num) %>% 
                              .$Geneid
                            #Get 1 for on the list and 0 for not.
                            gene_diff.list <- as.integer(gene.list %in% diff.list)
                            names(gene_diff.list) <- gene.list
                            return(gene_diff.list)
                          })
names(z_max_diff.list) <- paste0(seq(0,4)*3, 'h')

```{r phaseograms}
lapply(paste0(seq(0,4)*3, 'h'), 
       function(phase){
         draw_heatmap_zscore(zscore_list = chab_z.list,
                             filter_genes_vector = names(
                               getElement(z_max_diff.list, paste(phase))[
                                 getElement(z_max_diff.list, paste(phase)) == 1
                               ]
                             ),
                             gene_ordering_vector = order_z_rMT,
                             threshold_tpm = 1,
                             use_times = TRUE,
                             use_order_vector = FALSE,
                             show_row_names = FALSE,
                             tpm_mat = TRUE,
                             column_title_name = paste0('Genes ',phase, ' out-of-phase'))
         return()
       })
```

How many of these genes do not have GO annotations?

```{r phase_go_enrich}

@@

```{r}
#Load for the showSigOfNodes network function.

@@

#Get first (most significant) enriched GO id 
goID <- goRes[1, "GO.ID"]
#Show distribution of GO term genes compared to null distribution (all other genes)
print(showGroupDensity(GOdata, goID, ranks = TRUE))

@@

@@

GOres.list <- lapply(seq(0,4), 
                     function(phase){
                       GOdata <- new("topGOdata", 
                                     ontology = "BP", 
                                     geneSel = function(x){x == 1},
                                     allGenes = getElement(z_max_diff.list, paste(phase)),
                                     nodeSize = 5,
                                     annot = annFUN.gene2GO, 
                                     gene2GO = go_annotation)
                       resultFis <- runTest(GOdata, 
                                            algorithm = "classic", 
                                            statistic = "fisher")
                       gentable <- GenTable(GOdata, 
                                              classic = resultFis, 
                                              orderBy = "classic", 
                                              ranksOf = "classic",
                                              topNodes = 25) %>% 
                           filter(classic < 0.05)
                       return(list(GOdata = GOdata,
                                   resultFis = resultFis,
                                   gentable = gentable))
                     })
names(GOres.list) <- paste0(seq(0,4)*3, 'h')
lapply(GOres.list, getElement, 'gentable')

@@

```{r topGO_phase}
lapply(names(GOres.list), 
       function(GOres.num){
         GOres <- getElement(GOres.list, GOres.num)
         #Get first (most significant) enriched GO id 
         goID <- GOres$gentable[1, "GO.ID"]
         goTerm <- GOres$gentable[1, "Term"]

         #Show distribution of GO term genes compared to null distribution (all other genes)
         print(showGroupDensity(GOres$GOdata, goID, ranks = TRUE))

         #Can use genesInTerm function to extract the genes related to that GO term
         genesinterm <- genesInTerm(GOres$GOdata, goID)
         goID_genes <- unlist(genesinterm)
         names(goID_genes) <- genesinterm

         draw_heatmap_zscore(zscore_list = chab_z.list,
                             filter_genes_vector = goID_genes,
                             gene_ordering_vector = order_z_rMT,
                             threshold_tpm = 1,
                             use_times = TRUE,
                             use_order_vector = FALSE,
                             show_row_names = FALSE,
                             tpm_mat = TRUE,
                             column_title_name = paste0(GOres.num, ' - ', goTerm)
         )
       })

Note that some enriched GO terms may not appear in the annotation file (.gaf) directly as they are found to be enriched through the network of related/ancestral terms from the GO network of relations.

Also these phaseograms show every gene in entire dataset associated with the top GO term, so they may not all be as out-of-phase as one another. This is a good test to see if this GO enrichment seems correct - if most of the genes look like they're not in the phase expected then the GO enrichment is probably erroneous.
